Division and Factorization of p-Adic Polynomials

### 2016-01-30 polynomials/muldivfact.tex
Wrote an algorithm for slope factorization

@@

\begin{abstract}
We study two important operations on polynomials defined over complete 
discrete valuation fields: Euclidean division and factorization. In particular,
we design a simple and efficient algorithm for computing slope factorizations,
based on Newton iteration.  One of its main features is that we avoid working
over an extension of the base field.
We pay particular attention to stability, and analyze the behavior
of the algorithm using several precision models.
\end{abstract}

@@

Any computation with $p$-adic fields must work with approximations modulo finite
powers of $p$, and one of the key requirements in designing an algorithm is
an analysis of how the precision of the variables evolve over the computation.
We work with precision models developed by the same authors
\cite{caruso-roe-vaccon:14a}*{\S 4.2}, focusing on the lattice and Newton models.
As part of the analysis of the slope factorization algorithm, we describe
how the precision of the quotient and remainder depend on the input polynomials
in Euclidean division..

@@

\noindent
{\bf Main Results.}

Suppose that the Newton polygon of $P(X) = \sum_{i=0}^n a_i X^i$
has a break at $i=d$.  Set $A_0 = \sum_{i=0}^d a_i X^i$, $V_0 = 1$ and
\begin{align*}
A_{i+1} &= A_i + (V_i P \% A_i) \\
B_{i+1} &= P // A_{i+1} \\
V_{i+1} &= (2V_i - V_i^2B_{i+1}) \% A_{i+1}.
\end{align*}
Our main result is Theorem \ref{theo:slope-factor},
which states that the sequence $(A_i)$ converges quadritically
to a divisor of $P$.

@@

\noindent
{\bf Organization of the paper.}

After setting notation, in
Section \ref{sec:prec_data} we recall various models for tracking
precision in polynomial arithmetic.  We give some background
on Newton polygons and explain how using lattices to store
precision can allow for extra \emph{diffuse} $p$-adic digits that
are not localized on any single coefficient.

@@

Finally, in Section \ref{sec:slope_fac} we describe
our slope factorization algorithm, which is based on
a Newton iteration.  Unlike other algorithms for
slope factorization, ours does not require working
in an extension field \todo{citation?}.  In Theorem
\ref{theo:slope-factor} we define a sequence
of polynomials that will converge to the factors
determined by an extremal point in the Newton
polygon.  We then discuss the precision behavior
of the algorithm.

@@

If $\lambda_0 = -\infty$ or $\lambda_1 = +\infty$, the sequence $(A_i)$
is constant and the theorem is obvious. We then assume that $\lambda_0$
and $\lambda_1$ are both finite.
We notice that, when the polynomial is changed into $cP$ where $c$ is a 
nonzero constant lying in a finite extension of $K$, the $A_i$'s are all 
multiplied by $c$ as well whereas the $B_i$'s and the $V_i$'s remained 
unchanged. Therefore, the theorem holds for $P$ if and only if it holds 
for $cP$. As a consequence we may assume that $P$ is normalized so that 
$\NF(P)(d) = d \lambda_0$. We define the function $\varphi : \R^+ \to \R 
\cup \{+\infty\}$ by:
\begin{equation} \label{eqdef:functionphi}
\varphi(x) = \begin{cases} \NF(P)(x) &\mbox{ if $x \leq d$} \\
\lambda_0 (x-d) + \NF(P)(d) &\mbox{ if $x > d$} \end{cases}
%\begin{array}{r@{\,\,}ll}
%\varphi(x) & = \NP(P)(x) & \text{for } x \leq d \smallskip \\
%& = \lambda_0 (x-d) + \NP(P)(d) & \text{for } x > d.
%\end{array}
\end{equation}
\todo{For $x>d$, why not say $\varphi(x) = \lambda_0 x$?} For a polynomial $Q \in K[X]$ of degree $n$, we further define:
\begin{align}
b_\varphi(Q) & = \min_{x \in [0,n]} \NF(Q)(x) - \varphi(x) 
\label{eqdef:bphi} \\
\text{and} \quad
b_i(Q) & = \min_{x \in [0,n]} \NF(Q)(x) - \lambda_i x
\quad \text{for } i \in \{0,1\}.
\label{eqdef:b0}
\end{align}
Set also $b_\varphi(0) = b_0(0) = b_1(0) = +\infty$ by convention.
With the normalization of $P$ we chose above, we have $b_0(P)
= b_\varphi(P) = 0$ and $b_\varphi(Q) \leq b_0(Q)$ for all polynomial
$Q$. Similarly $b_1(Q) \leq b_0(Q)$ for all $Q$.

@@

Moreover, the conclusion of 
Theorem~\ref{theo:slope-factor} is still correct if $A_0$ is any 
polynomial of degree $d$ and leading coefficient $a_d$ and $V_0$
is any polynomial as soon as they satisfy:
$$b_\varphi\big(V_0 P \mod A_0\big) > 0
\quad \text{and} \quad
b_0\big((1 - V_0 B_0) \mod A_0\big) > 0$$
except that the constant $\kappa$ giving the rate of convergence
should be now defined as:
$$\kappa = \min \big( 
b_\varphi\big(V_0 P \mod A_0\big), \,
b_0\big((1 - V_0 B_0) \mod A_0\big)\big).$$
\end{rem}

@@

\todo{Answer here the second question.}

@@

\todo{Answer here the second question. (The Proposition below should
be useful at some point.)}

@@

\bibliographystyle{plain}
\bibliography{../roebib/Biblio}

>>>>

\newcommand{\softO}{O\tilde{~}}

@@

\begin{abstract}
We study two important operations on polynomials defined over complete 
discrete valuation fields: Euclidean division and factorization. In particular,
we design a simple and efficient algorithm for computing slope factorizations,
based on Newton iteration.  One of its main features is that we avoid working
with fractional exponents.
We pay particular attention to stability, and analyze the behavior
of the algorithm using several precision models.
\end{abstract}

@@

Any computation with $p$-adic fields must work with approximations modulo finite
powers of $p$, and one of the key requirements in designing an algorithm is
an analysis of how the precision of the variables evolve over the computation.
We work with precision models developed by the same authors
\cite{caruso-roe-vaccon:14a}*{\S 4.2}, focusing on the lattice and Newton models.
As part of the analysis of the slope factorization algorithm, we describe
how the precision of the quotient and remainder depend on the input polynomials
in Euclidean division.

@@

\noindent
{\bf Main Results.}
%
Suppose that the Newton polygon of $P(X) = \sum_{i=0}^n a_i X^i$
has a break at $i=d$.  Set $A_0 = \sum_{i=0}^d a_i X^i$, $V_0 = 1$ and
\begin{align*}
A_{i+1} &= A_i + (V_i P \% A_i) \\
B_{i+1} &= P \div A_{i+1} \\
V_{i+1} &= (2V_i - V_i^2B_{i+1}) \% A_{i+1}.
\end{align*}
Our main result is Theorem \ref{theo:slope-factor},
which states that the sequence $(A_i)$ converges quadritically
to a divisor of $P$.

@@

\noindent
{\bf Organization of the paper.}
%
After setting notation, in
Section \ref{sec:prec_data} we recall various models for tracking
precision in polynomial arithmetic.  We give some background
on Newton polygons and explain how using lattices to store
precision can allow for extra \emph{diffuse} $p$-adic digits that
are not localized on any single coefficient.

@@

Finally, in Section \ref{sec:slope_fac} we describe
our slope factorization algorithm, which is based on
a Newton iteration.  Unlike other algorithms for
slope factorization, ours does not require working
with fractional exponents \todo{citation?}.  In Theorem
\ref{theo:slope-factor} we define a sequence
of polynomials that will converge to the factors
determined by an extremal point in the Newton
polygon.  We then discuss the precision behavior
of the algorithm.

@@

If $\lambda_0 = -\infty$ or $\lambda_1 = +\infty$, the sequence $(A_i)$
is constant and the theorem is obvious. We then assume that $\lambda_0$
and $\lambda_1$ are both finite.

We define the function $\varphi : \R^+ \to \R 
\cup \{+\infty\}$ by:
\begin{equation} \label{eqdef:functionphi}
\varphi(x) = \begin{cases} \NF(P)(x) &\mbox{ if $x \leq d$} \\
\lambda_0 (x - d) + \NF(P)(d) &\mbox{ if $x > d$} \end{cases}
%\begin{array}{r@{\,\,}ll}
%\varphi(x) & = \NP(P)(x) & \text{for } x \leq d \smallskip \\
%& = \lambda_0 (x-d) + \NP(P)(d) & \text{for } x > d.
%\end{array}
\end{equation}
We notice that, when the polynomial is changed into $cP$ where $c$ is a 
nonzero constant lying in a finite extension of $K$, the $A_i$'s are all 
multiplied by $c$ as well whereas the $B_i$'s and the $V_i$'s remained 
unchanged. Therefore, the theorem holds for $P$ if and only if it holds 
for $cP$. As a consequence we may assume that $P$ is normalized so that 
$\NF(P)(d) = d \lambda_0$, so that $\varphi(x) = \lambda_0 x$ for $x > d$.
For a polynomial $Q \in K[X]$ of degree $n$, we further define:
\begin{align}
b_\varphi(Q) & = \min_{x \in [0,n]} \NF(Q)(x) - \varphi(x) 
\label{eqdef:bphi} \\
\text{and} \quad
b_i(Q) & = \min_{x \in [0,n]} \NF(Q)(x) - \lambda_i x
\quad \text{for } i \in \{0,1\}.
\label{eqdef:b0}
\end{align}
Set also $b_\varphi(0) = b_0(0) = b_1(0) = +\infty$ by convention.
With the normalization of $P$ we chose above, we have $b_0(P)
= b_\varphi(P) = 0$ and $b_\varphi(Q) \leq b_0(Q)$ for all polynomial
$Q$. Similarly $b_1(Q) \leq b_0(Q)$ for all $Q$.

@@

Moreover, the conclusion of 
Theorem~\ref{theo:slope-factor} is still correct if $A_0$ is any 
polynomial of degree $d$ with leading coefficient $a_d$ and $V_0$
is any polynomial as soon as they satisfy:
$$b_\varphi\big(V_0 P \mod A_0\big) > 0
\quad \text{and} \quad
b_0\big((1 - V_0 B_0) \mod A_0\big) > 0$$
except that the constant $\kappa$ giving the rate of convergence
should be now defined as:
$$\kappa = \min \big( 
b_\varphi\big(V_0 P \mod A_0\big), \,
b_0\big((1 - V_0 B_0) \mod A_0\big)\big).$$
\end{rem}

@@

We can now move to the second question we have raised before, 
\emph{i.e.} the design of an algorithm for computing $A^{(1)}$ with the 
precision given by Proposition~\ref{prop:Newtonprecslope}. Our strategy 
consists in computing first the precision and applying then the Newton 
iteration until the expected precision is reached. Below is the
precise description of our algorithm.

\noindent {\bf Algorithm} {\tt slope\_factorisation\_Newton}

\noindent{\bf Input:} a monic polynomial $P + O(\varphi_P) \in K_n[X]$,

\noindent\phantom{{\bf Input:} }a break point $d$ of $\NP(P)$

\noindent{\bf Output:} the factor $A$ described above

\smallskip\noindent 1.\ %
Compute the functions $\NF(P)$ and $\varphi$

\smallskip\noindent 2.\ %
Compute $\varphi_A = \varphi_{|[0,d-1]} - \varphi(d) + \min_{x \in [0,n]}
\varphi_P(x) - \varphi(x)$

\smallskip\noindent 3.\ %
Compute $\kappa = \NP(P)(d{+}1) + \NF(P)(d{-}1) - 2 \cdot \NF(P)(d)$

\smallskip\noindent 4.\ %
Set $i = 0$, $A_0 = \sum_{i=0}^d a_i X^i$
($a_i = \text{coeffs of } P$), $V_0 = 1$

\smallskip\noindent 5.\ %
{\bf repeat until} $\varphi_{|[0,d{-}1]} + 2^i \kappa \geq \varphi_A$

\smallskip\noindent 6.\ \hspace{5mm}%
lift $A_i$, $V_i$ and $P$ at enough precision

\smallskip\noindent 7.\ \hspace{5mm}%
compute

\hspace{1cm}$\bullet$
$A_{i+1} = A_i + (P V_i \mod A_i)$

\hspace{1cm}\phantom{$\bullet$ }%
at precision $O(\varphi_{|[0,d{-}1]} + 2^{i+1} \kappa)$

\hspace{1cm}$\bullet$
$V_{i+1} = (2 V_i - V_i^2 \cdot (P \div A_{i+1})) \mod A_{i+1}$

\hspace{1cm}\phantom{$\bullet$ }%
at precision $O(x \mapsto \lambda_1 x + 2^{i+1}\kappa)$

\smallskip\noindent 8.\ \hspace{5mm}%
set $i = i+1$

\smallskip\noindent 9.\ 
{\bf return} $A_i$

\begin{rem}
The precision needed at line 6 is of course governed by the computation
performed at line 7. Note that it can be either computed \emph{a
priori} by using Proposition~\ref{prop:NewtonprecEuclide} or dynamically 
by using relaxed algorithms from 
\cite{hoeven:02a,hoeven:07a,berthomieu-hoeven-lecerf:11a}. In both 
cases, it is in $O(\pi^{N_i})$ with $N_i = O(2^i \kappa + \min 
\NF(P))$.
\end{rem}

\noindent
It follows from Theorem~\ref{theo:slope-factor}, 
Remark~\ref{rem:slope-factor} and Proposition~\ref{prop:Newtonprecslope} 
that Algorithm {\tt slope\_factorisation\_Newton} is correct and 
stable. Using the standard soft-$O$ notation $\softO(\cdot)$ for
hiding logarithmic factor, our algorithm performs at most $\softO(n)$ 
combinatorial operations and $\softO(n)$ operations in $K$ at precision 
$O(\pi^N)$ with $N = O(\max \varphi_P - \min \NF(P))$ if one uses 
quasi-optimal algorithms for multiplication and Euclidean division of 
polynomials.

@@

As for the second question, the discussion is similar to the case of 
Newton precision expect that we need a new stop criterium. It is given 
by the following proposition.

@@

\noindent
{\bf Numerical experimentations.} 
%
We have implemented the following scheme to illustrate the behaviour of 
precision for slope factorization over $\mathbb{Q}_2 [X].$
We start with random polynomials $A$ and $B$ with given minoration on 
their Newton polygons, assuring they have no common slope.
We compute $P=AB$ and then $A^{(1)}$, and compare the obtained 
precision on $A^{(1)}$ (jagged) with the estimation given using 
$\psi_{A^{(1)}}$ and Proposition \ref{prop:precA1} 
assuming~\textbf{(H)}. The mean amount of lost digits for $1000$ 
experiments is given in the following array. Two settings are possible 
for the initial $A$ and $B$ : relative or absolute precision equal to 
$100$ for their coefficients.

{\small%
\noindent \hfill%
\renewcommand{\arraystretch}{1.3}%
\begin{tabular}{|c|c|c|c|c|}%
\hline
\multirow{2}{*}{Polynomials} & 
\multirow{2}{*}{\hspace{1mm}Precision\hspace{1mm}} & 
\multicolumn{3}{c|}{Gain of precision} \\
\cline{3-5}
& &Jagged & Newton & Lattice \\
\hline
\multirow{2}{*}{%
\begin{tikzpicture}[xscale=0.11, yscale=0.05]
\draw[transparent] (-2,11) rectangle (15,11);
\draw[->] (0,-.5) -- (0,10.5);
\draw[->] (-0.05,0) -- (13.5,0);
\draw (6,-.75) -- (6,.75);
\node[scale=0.7,below] at (6,0) {$d$};
\draw (0,5) -- (2,-2) -- (6,0) -- (10,3) -- (13,10);
\draw[dotted] (6,0) -- (6,10);
\node at (3.5,5) {$A$};
\node at (8.5,5) {$B$};
\end{tikzpicture}}
& absolute & $-14.5$ & $-14.7$ & $0.0$ \\
& relative & $\phantom{0}{-}1.5$ & $\phantom{0}{-}3.6$ & $0.0$ \\
\hline
\multirow{2}{*}{\begin{tikzpicture}[xscale=0.11, yscale=0.05]
\draw[transparent] (-2,11) rectangle (15,11);
\draw[->] (0,-.5) -- (0,10.5);
\draw[->] (-0.05,0) -- (13.5,0);
\draw (6,-.75) -- (6,.75);
\node[scale=0.7,below] at (6,0) {$d$};
\draw (0,5) -- (6,0) -- (11,-2) -- (13,10);
\draw[dotted] (6,0) -- (6,10);
\node at (3.5,5) {$A$};
\node at (8.5,5) {$B$};
\end{tikzpicture}}
& absolute & $\phantom{-0}0.0$ & $\phantom{-0}0.0$ & $0.0$ \\
& relative & $\phantom{0}{-}0.3$ & $\phantom{0}{-}1.2$ & $0.0$ \\
\hline
\end{tabular}
\hfill \null}

\bibliographystyle{plain}
\bibliography{muldivfact}
